Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-03-28 GWLDCI.tex
typos in sections 1 and 2.1, and clarified things in the associahedron proof as discussed with Susama yesterday

In recent years, there has been a an active program researching the geometric and combinatorics underlying SYM $N=4$ theory \cite{wilsonloops, amplituhedron, correlahedron, karpwilliams, galashinlam}. This body of work started with the observation that the on shell amplitudes of this theory correspond to the volume of a subspace of a positive Grasmannian, called an Amplituhedron \cite{amplituhedron}. Since then, there has been significant work to study the structure of the amplituhedron, geometrically and combinatorially, for instance \cite{amplituhedron2, bigamplituhedronpaper, karpwilliams, galashinlam}. The focus of much of this work has been to understand the relationship between the combinatorial objects from the physics of the system, BCFW diagrams, and their corresponding natural object for studying the positive Grassmannians, positroid cells.

Meanwhile, there has been a different body of work studying SYM $N=4$ theory from the point of view of Feynman integrals in twistor space\cite{BullimoreSkinner, AldayMaldecena?, Wittenpaper}. These integrasl are calculated via Wilson loops. The first author and Amat uncovered a connection between the Feynman integrals developed in this literature and positroids, defined  as matroids that can be realized as an element of $\Gr(k,n)$ \cite{wilsonloops}. With the second author and others, this work was extended to study these Feynman integrals in terms of positroid cells and the positive Grassmanians \cite{casestudy, non-orientable}. Others have tried different approaches to define a geometric space associated to these integrals in a manner similar to the Amplituhedron \cite{correlahedron, HeslopStewart}.

@@

We proceed in Section~\ref{section overall background} with the required definitions for Wilson loop diagrams (subsection~\ref{section WLD background}) and matroids (subsection~\ref{sec matroid background}.  Wilson loop diagrams  are said to be admissible when they correspond to $N^kMHV$ diagrams, or mathematically according to Definition~\ref{admisdfn}.  With this, the key definition is Definition~\ref{equivdfn} giving the equivalence relation on admissible Wilson loop diagrams.  The connection to matroids is stated precisely in Theorem \ref{thm WLD defines matroid}.

@@

We depict this data as a circle with marked points, called vertices. The vertices are labeled by $V$ (preserving the cyclic ordering). The arc between consecutive vertices are called edges. There are $k$ wavy lines in the interior of the diagram, depicting the proagators, with endpoints on the edges. A propagator, $p =(i,j)$ has one endpoint on the edge between the vertex labeled $i$ and $i+1$ and another endpoint on the edge defined by $j$ and $j+1$. The condition on $i_r$ and $j_r$ means that the propagator does not go between adjacent edges. Let $\cP = \{p_r\}_{r=1}^k$ be the set of propagators. Then we write \bas W = (\cP, V) \;.\eas \sanote{picture}

@@

\begin{rmk}\label{alt F(P) rmk}
Note that a more explicit definition of $F(P)$ is \bas F(P)  = \big(V(P) \setminus V(P^c) \big) \cup F(\emptyset) \;.\eas Furthermore, note that by construction $\Prop(F(P)) \subseteq P$.  
\end{rmk}

@@

Note that if we take any admissible Wilson loop diagram and remove the unsupported vertices then we will obtain a weakly admissible Wilson loop diagram that may or may not be admissible itself.

@@

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note that since vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$, $M(W)$ can also be thought of as the matroid realized by $C(W)$.

@@

@@

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{}; the following will be the one which we will use. \sanote{picture of an associahedron somewhere? or do people know what it is?}

@@

\begin{proof}
  Let $W$ and $X$ be two equivalent Wilson loop diagrams.  From \hlfix{Lemma {\ref{??}}}{Do we actually have this explicitly? Feels like we should}, we know that $\tau(W)$ and $\tau(X)$ differ from each other only by retriangulations of their triangulated pieces.

@@

  Returning to $\tau(W)$ and $\tau(X)$ since the direction vector of an edge in $A_n$ depends only on its quadrilateral and the bijection between the edges bounding $\tau(W)$ and the edges bounding $\tau(X)$ preserves the quadrilateral, we have the same direction vectors bounding $\tau(W)$ as $\tau(X)$.  Furthermore, the bijection also maintains the relative configuration of the edges and so $\tau(W)$ and $\tau(X)$ are parallel.
\end{proof}

The following lemma is useful for proving the converse of the above.

\begin{lem}\label{lem good quads}
  Let $D_1$ and $D_2$ be two dissections of the convex polygon defined by the vertices $x_1, \ldots, x_n$.  \hlfix{Suppose the vertices incident to at least one non-triangle face are the same in both dissections}{relevant?}, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ such that in one of $D_1$ or $D_2$ all four vertices are on the same non-triangle face while in the other there is a diagonal of the dissection such that there is one side of the diagonal with exactly one of $\{x_i, x_j, x_k, x_l\}$ (so the others are end points of this diagonal or strictly on the other side).
\end{lem} \sanote{picture?}

@@

Suppose this never occurs.  Then all the non-triangle faces of $D_2$ which are crossed by a diagonal of $D_1$ must be quadrilaterals and likewise for $D_1$.  This means that any non-triangle and non-quadrilateral faces must be common between $D_1$ and $D_2$.  \hlfix{These common faces are irrelevant for the statement}{then why in the statement of thm?}, and so by triangulating any common non-triangle faces, we may assume that $D_1$ and $D_2$ have only quadrilateral and triangle faces.

Assume for a contradiction that no quadrilateral as in the statement exists, and let $d$ and $f$ be as above.  Then $f$ must have two vertices strictly on each side of $d$, say $x_i$ and $x_j$ on one side and $x_k$, $x_l$ on the other side (and in that order).  Consider the face incident to $d$ on the $x_i, x_j$ side of $d$.  Suppose this face is a triangle, then its other vertex cannot be in $[x_i, x_j]$ as otherwise taking one of the other edges of the triangles would give a configuration as in the statement.  Suppose now that this face is a quadrilateral, again no other vertex of this face can be in $(x_i, x_j)$ as otherwise again taking one of the other edges of the face as $d$ we'd get a quadrilateral as in the statement, the same holds if one of the other vertices of the face is $x_i$ or $x_j$ but the other is not.  If the two other vertices of the face are $x_i$ and $x_j$, then we also get a configuration as in the statement but now with the four vertices being the two ende of $q$ and $x_i, x_j$ and with one of the edges of $f$ as the diagonal.  Thus we can conclude that none of the vertices of this face are in $[x_i, x_j]$.  Also if the face has three vertices on one side of $f$ and one on the other then we get a quadrilateral as in the statement again with this face giving the four vertices and the diagonal a side of $f$.  , so the face is either a triangle with its third vertex of the triangle must be in $(x_j,x_k)$ or in $(x_l,x_i)$ or a quadrilateral with one more vertex in $(x_j, x_k)$ and one more vertex in $(x_l, x_i)$.  Taking all this together, whether the face is a triangle or a quadrilateral, replacing $d$ with one of the other edges of the face we get a new diagonal $d'$ which also separates $f$ into $x_i, x_j$ and $x_k, x_l$ but is strictly closer to $[x_i, x_j]$ than $d$ was.  The same argument holds for $d'$, but the polygon is finite, so this process cannot go on indefinitely, contradicting that no quadtilateral as in the statement exists.
\end{proof}

@@

\begin{proof}
  Let $W$ and $X$ be inequivalent.  If $\tau(W)$ and $\tau(X)$ do not have the same number of diagonals (hence their faces in $A_n$ having the same dimension) then they are not parallel, so we can assume $\tau(W)$ and $\tau(X)$ do have the same number of diagonals and hence their faces in $A_n$ have the same dimension.  Let the face associated to $\tau(W)$ be $f_W$ and the face associated to $\tau(X)$ be $f_X$.

  As observed in the proof of Proposition~\ref{prop easy way}, the support of any nonzero vector parallel to an edge is on the indices corresponding to the four vertices bounding the quadrilateral of the dissection associated with the edge.  Thus $\tau(W)$ and $\tau(X)$ do not have the same set of vertices incident to at least one non-triangle face then the span of the vectors parallel to their edes has different support and so they cannot be parallel.  So we can assume that $\tau(W)$ and $\tau(X)$ have the same set of vertices incident to at least one non-triangle face.

    Since $\tau(W)$ and $\tau(X)$ do not have the same set of non-triangle faces, there must be some diagonal $d$ of $\tau(W)$ that crosses through a non-triangle face $f$ of $\tau(X)$. 
    It will be most convenient to look at normal vectors and to use the choice of normal vector discussed in the many realizations paper on p8, namely pick one side of $d$ to be positive and define the $i$th coordinate of the $\omega^+_d$ to be
    \[
    \omega^+_d(i) = \begin{cases} \text{dist}(x_i, d) & \text{if $x$ is on the positive side of $d$}\\ 0 & \text{otherwise} \end{cases}
    \]
    The many realizations paper explains why $\omega^+_d$ is a normal vector for any face with $d$ as a diagonal in a way which is short and clean but relies on some earlier definitions.  We can instead check that it is a normal vector in a completely elementary way using the shoelace formula for the area of a triangle which says that if $(x_1, y_1), (x_2, y_2), (x_3, y_3)$ are the coordinates of the corners of a triangle in counterclockwise order then the area of the triangle is $(x_1y_2 - x_1y_3 + x_2y_3 - x_2y_1 + x_3y_1 - x_3y_2)/2$.

>>>>

In recent years, there has been a an active program researching the geometry and combinatorics underlying SYM $N=4$ theory \cite{wilsonloops, amplituhedron, correlahedron, karpwilliams, galashinlam}. This body of work started with the observation that the on shell amplitudes of this theory correspond to the volume of a subspace of a positive Grasmannian, called an Amplituhedron \cite{amplituhedron}. Since then, there has been significant work to study the structure of the amplituhedron, geometrically and combinatorially, for instance \cite{amplituhedron2, bigamplituhedronpaper, karpwilliams, galashinlam}. The focus of much of this work has been to understand the relationship between the combinatorial objects from the physics of the system, BCFW diagrams, and their corresponding natural object for studying the positive Grassmannians, positroid cells.

Meanwhile, there has been a different body of work studying SYM $N=4$ theory from the point of view of Feynman integrals in twistor space\cite{BullimoreSkinner, AldayMaldecena?, Wittenpaper}. These integrals are calculated via Wilson loops. The first author and Amat uncovered a connection between the Feynman integrals developed in this literature and positroids, defined  as matroids that can be realized as an element of $\Gr(k,n)$ \cite{wilsonloops}. With the second author and others, this work was extended to study these Feynman integrals in terms of positroid cells and the positive Grassmanians \cite{casestudy, non-orientable}. Others have tried different approaches to define a geometric space associated to these integrals in a manner similar to the Amplituhedron \cite{correlahedron, HeslopStewart}.

@@

We proceed in Section~\ref{section overall background} with the required definitions for Wilson loop diagrams (subsection~\ref{section WLD background}) and matroids (subsection~\ref{sec matroid background}).  Wilson loop diagrams  are said to be admissible when they correspond to $N^kMHV$ diagrams, or mathematically according to Definition~\ref{admisdfn}.  With this, the key definition is Definition~\ref{equivdfn} giving the equivalence relation on admissible Wilson loop diagrams.  The connection to matroids is stated precisely in Theorem \ref{thm WLD defines matroid}.

@@

We depict this data as a circle with marked points, called vertices. The vertices are labeled by $V$ (preserving the cyclic ordering). The arc between consecutive vertices are called edges. There are $k$ wavy lines in the interior of the diagram, depicting the proagators, with endpoints on the edges. A propagator, $p =(i,j)$ has one endpoint on the edge between the vertex labeled $i$ and $i+1$ and another endpoint on the edge defined by $j$ and $j+1$. The condition\note{which condition is this referring to; I think it is talking about one of the admissibility conditions, so maybe this is in the wrong place, alternately the condition might have gotten deleted at some point} on $i_r$ and $j_r$ means that the propagator does not go between adjacent edges. Let $\cP = \{p_r\}_{r=1}^k$ be the set of propagators. Then we write \bas W = (\cP, V) \;.\eas \sanote{picture}

@@

\begin{rmk}\label{alt F(P) rmk}
Note that a more explicit definition of $F(P)$ is \bas F(P)  = \big(V(P) \setminus V(P^c) \big) \cup F(\emptyset) \;.\eas Furthermore, note that by construction $\Prop(F(P)) \subseteq P$.  This containment is strict if there is any propagator $p$ in $P$ with the property that every vertex in the support of $p$ also supports a propagator not in $P$.
\end{rmk}

@@

Note that if we take any admissible Wilson loop diagram and remove the non-supporting vertices then we will obtain a weakly admissible Wilson loop diagram that may or may not be admissible itself.

@@

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note that since vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$, $M(W)$ can also be thought of as the matroid realized by $C(W)$. \note{Can this note have a ref to your earlier paper?  If not then perhaps expand slightly as although I agree that it is true, I don't think it is quite so trivial.  Also, I think ``is also'' woulc be fine in place of ``can also be thought of as''}

@@

It will be important in the subsequent section to consider retriangulations.
\begin{dfn}
  If $W$ and $W'$ are weakly admissible Wilson loop diagrams on the same vertex set then $\tau(W)$ and $\tau(W')$  differ by a \emph{retriangulation} if there is a bijection $f$ between the maximal triangulated pieces of $\tau(W)$ and the maximal triangulated pieces of $\tau(W')$ such that for any triangulated piece $t$ of $\tau(W)$ the vertex sets of $t$ and $f(t)$ are the same.
\end{dfn}
More casually this definition says that the triangulated pieces of $\tau(W)$ and $\tau(W')$ are the same (so far as their vertex sets are concerned) but that some or all of the triangulated pieces may be differently triangulated while the region being triangulated by each piece (which is defined by the vertex set) remains unchanged.

The previous definition makes sense, in fact, for any two polygon dissections of the same polygon.

\medskip

@@

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{}; the following will be the one which we will use. \sanote{picture of an associahedron somewhere?}\note{Yes, I agree that there should be a picture}

@@

\begin{proof}
  Let $W$ and $X$ be two equivalent Wilson loop diagrams.  From Lemma~\ref{lem triang to exact} along with the definitions of $\sim$ and retriangulation, we know that $\tau(W)$ and $\tau(X)$ differ from each other only by retriangulations of their triangulated pieces.

@@

  Returning to $\tau(W)$ and $\tau(X)$, since the direction vector of an edge in $A_n$ depends only on its quadrilateral and the bijection between the edges bounding $\tau(W)$ and the edges bounding $\tau(X)$ preserves the quadrilateral, we have the same direction vectors bounding $\tau(W)$ as $\tau(X)$.  Furthermore, the bijection also maintains the relative configuration of the edges and so $\tau(W)$ and $\tau(X)$ are parallel.
\end{proof}

The following lemma is useful for proving the converse of the above. We also need the following notation.  Given a polygon dissection $D$ let $V_{>3}(D)$ be the set of vertices of the polygon which are on at least one non-triangle face of $D$.  Note that $D$ may have many non-triangle faces and so $V_{>3}(D)$ will be the union of the vertices of each non-triangle face.

\begin{lem}\label{lem good quads}
  Let $D_1$ and $D_2$ be two dissections of the convex polygon defined by the vertices $x_1, \ldots, x_n$.  Suppose $V_{>3}(D_1) = V_{>3}(D_2)$, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ such that in one of $D_1$ or $D_2$ all four vertices are on the same non-triangle face while in the other there is a diagonal of the dissection such that there is one side of the diagonal with exactly one of $\{x_i, x_j, x_k, x_l\}$ (so the others are end points of this diagonal or strictly on the other side).
\end{lem} \sanote{picture?}

@@

Suppose this never occurs.  Then all the non-triangle faces of $D_2$ which are crossed by a diagonal of $D_1$ must be quadrilaterals and likewise for $D_1$.  This means that any non-triangle and non-quadrilateral faces must be common between $D_1$ and $D_2$.  These common faces are irrelevant for the statement as none of them can be the face containing $x_i, x_j, x_k, x_l$, and so by triangulating any common non-triangle faces, we may assume that $D_1$ and $D_2$ have only quadrilateral and triangle faces.

Assume for a contradiction that no quadrilateral as in the statement exists, and let $d$ and $f$ be as above.  Then $f$ must have two vertices strictly on each side of $d$, say $x_i$ and $x_j$ on one side and $x_k$, $x_l$ on the other side (and in that order).  Consider the face incident to $d$ on the $x_i, x_j$ side of $d$.  Suppose this face is a triangle, then its other vertex cannot be in $[x_i, x_j]$ as otherwise taking one of the other edges of the triangles would give a configuration as in the statement.  Suppose now that this face is a quadrilateral, again no other vertex of this face can be in $(x_i, x_j)$ as otherwise again taking one of the other edges of the face as $d$ we'd get a quadrilateral as in the statement, the same holds if one of the other vertices of the face is $x_i$ or $x_j$ but the other is not.  If the two other vertices of the face are $x_i$ and $x_j$, then we also get a configuration as in the statement but now with the four vertices being the two ende of $q$ and $x_i, x_j$ and with one of the edges of $f$ as the diagonal.  Thus we can conclude that none of the vertices of this face are in $[x_i, x_j]$.  Also if the face has three vertices on one side of $f$ and one on the other then we get a quadrilateral as in the statement again with this face giving the four vertices and the diagonal is a side of $f$.  In this case the face is either a triangle with its third vertex of the triangle must be in $(x_j,x_k)$ or in $(x_l,x_i)$ or a quadrilateral with one more vertex in $(x_j, x_k)$ and one more vertex in $(x_l, x_i)$.  Taking all this together, whether the face is a triangle or a quadrilateral, replacing $d$ with one of the other edges of the face we get a new diagonal $d'$ which also separates $f$ into $x_i, x_j$ and $x_k, x_l$ but is strictly closer to $[x_i, x_j]$ than $d$ was.  The same argument holds for $d'$, but the polygon is finite, so this process cannot go on indefinitely, contradicting that no quadtilateral as in the statement exists.
\note{That huge paragraph also needs a picture}
\end{proof}

@@

\begin{proof}
  Let $W$ and $X$ be inequivalent.  If $\tau(W)$ and $\tau(X)$ do not have the same number of diagonals (hence their faces in $A_n$ not having the same dimension) then they are not parallel, so we can assume $\tau(W)$ and $\tau(X)$ do have the same number of diagonals and hence their faces in $A_n$ have the same dimension.  Let the face associated to $\tau(W)$ be $f_W$ and the face associated to $\tau(X)$ be $f_X$.

  As observed in the proof of Proposition~\ref{prop easy way}, the support of any nonzero vector parallel to an edge is on the indices corresponding to the four vertices bounding the quadrilateral of the dissection associated with the edge.  Thus $V_{>3}(\tau(W)) \neq V_{>3}(\tau(X))$ (notation as defined before Lemma~\ref{lem good quads}) then the span of the vectors parallel to their edes has different support and so they cannot be parallel.  So we can assume that $V_{>3}(\tau(W)) = V_{>3}(\tau(X))$.

    Since $\tau(W)$ and $\tau(X)$ do not have the same set of non-triangle faces, there must be some diagonal $d$ of $\tau(W)$ that crosses through a non-triangle face $f$ of $\tau(X)$. 
    It will be most convenient to look at normal vectors and to use the choice of normal vector discussed on p8 of \cite{CSZinequivalent}, namely pick one side of $d$ to be positive and define the $i$th coordinate of the $\omega^+_d$ to be
    \[
    \omega^+_d(i) = \begin{cases} \text{dist}(x_i, d) & \text{if $x$ is on the positive side of $d$}\\ 0 & \text{otherwise} \end{cases}
    \]
    We have from \cite{CSZinequivalent} that $\omega^+_d$ is a normal vector for any face with $d$ as a diagonal, however we can also check this fact in a completely elementary way using the shoelace formula for the area of a triangle.  The shoelace formula says that if $(x_1, y_1), (x_2, y_2), (x_3, y_3)$ are the coordinates of the corners of a triangle in counterclockwise order then the area of the triangle is $(x_1y_2 - x_1y_3 + x_2y_3 - x_2y_1 + x_3y_1 - x_3y_2)/2$.
